ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-27 pyked.tex
Update and reflow intro

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding
fuel chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles.
However, despite both the importance and abundance of such information in the literature,
the combustion\slash chemical kinetics community lacks an accepted, commonly used
standard for recording and sharing data from fundamental combustion experiments.\footnote{
Interestingly, this contrasts the situation for chemical kinetic models, where the
CHEMKIN~\autocite{Kee:1996ck} format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code}
lag behind considerably, although use of the former continually grows due to Cantera's
open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?
}

Instead, such data is typically found in under-documented
comma-separated value files and Excel spreadsheets, or contained in PDF tables
rather than as supplementary material associated with a paper.
In the worst case, data is only available in figures and must be digitized by the
use of software such as WebPlotDigitizer~\autocite{WebPlotDigitizer}.
% I think we're already stirring the pot enough, no need to call out
% specific people
% \todo{call out specific examples of these...?}{}
These practices limit the wide use of the valuable, extensive experimental data
available in the literature. Frenklach~\autocite{Frenklach:2007bm} further explained the
benefits of standardized, widely available data for combustion.
In brief, much of combustion\slash chemical kinetics research drives towards the ultimate goal
of developing predictive kinetic models---and ultimately this depends on a community
infrastructure for data and methods.

The combustion community has not yet widely adopted a standard format for experimental data,
but some researchers and groups attempted to solve this problem.\todo[author=Kyle]{this still doesn't read quite right}{}
Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess \textbf{I}nformatics
\textbf{M}od\textbf{e}l) data format~\autocite{Frenklach:2007bm,You:2011hy} and an associated
online database, the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated quantities,
using the Extensible Markup Language (XML) standard.

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make it
a strong standardization format, but suffers from two primary flaws that
prevent wide adoption: (1) the PrIMe standard does not require or support all information
needed to simulate an experiment, including a machine-readable definition of ignition or a
standard way to express the rapid compression machine (RCM) compression process; (2) and
the XML format, intended more as a markup language than data format, presents a barrier
to creating and working with files.
Furthermore, the focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for scholarly products
and InCHI or SMILES for species), limits usability outside the PrIMe ecosystem.
Finally, the closed and opaque nature of the PrIMe standard and associated Data Warehouse
limit contributions and community development of these and compatible tools.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the ReSpecTh standard,
an XML-based format that builds on PrIMe. ReSpecTh adds important features that make files
better standalone representations of experimental data---i.e., more informative by themselves
rather than in concert with a larger system.
For example, where PrIMe used internal bibliographic references, ReSpecTh added a field for
typical bibliographic data. The ReSpecTh standard also provides machine-readable formats
for better specifying ignition experiments, including a definition of autoignition and
either volume histories for RCMs or pressure-rise in shock tubes.
However, being XML-based, ReSpecTh experimental files still suffer from the same usability
issues as PrIMe, and the standard remains closed.
Olm and colleagues used their ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models, and more recently
develop improved models for hydrogen\slash syngas~\autocite{Varga:2016gj} and
ethanol~\autocite{Olm:2016et}. In all four cases, they converted numerous experimental datasets
from the literature into the ReSpecTh standard.
\todo[author=Kyle,inline]{how's this?}

Building on the lack of an accepted data standard, limited examples exist of open sharing
of such data. The Clean Combustion Research Center at King Abdullah University of Science
and Technology (KAUST) started the CloudFlame project in 2013~\autocite{Goteng:2013cf}\todo{a few more refs needed}{}
(\url{https://cloudflame.kaust.edu.sa}). CloudFlame does serve as an openly accessible database
for experimental data, available in a standard CSV format, but also goes beyond this by
providing a cloud infrastructure for running simulations based on stored models and data.
While an admirable (and useful) effort, CloudFlame represents another closed system controlled
by a specific institution rather than the community.

In this work, we present a new, open-source, human- and machine-readable data standard for fundamental combustion
experiments, \ck{}, and offer tools for easily working with data encoded in this format.
Niemeyer recently introduced an initial version of the \ck{} format~\autocite{Niemeyer:2016wf},
and in this work we further formalize and develop the standard. We also discuss a related
Python software tool for validating and working with \ck{} files: \pk{}.
Our motivations are similar those of the PrIMe and ReSpecTh teams:
the \ck{} project will enable easy sharing and use of fundamental combustion data,
for the (primary) purposes of developing and validating predictive chemical kinetic models.
However, we make usability a major design focus, and plan to share all data and software openly
to cultivate a community of user-contributors.

@@

\ck{} and \pk{} are currently under active development, and contributions from
the community are gratefully welcomed. All development occurs under the BSD
3-clause license and the code is housed on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}. Future directions for development
are outlined in the public
\href{Roadmap}{https://github.com/pr-omethe-us/PyKED/wiki/Roadmap}. The highest
priority issues are currently adding support for other types of fundamental
experiments, including flame speed, speciation, and flame extinction
measurements.

>>>>

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding fuel
chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles,
among others. However, despite both the importance and abundance of such
information in the literature, the combustion\slash chemical kinetics community
lacks an accepted, commonly used standard for recording and sharing data from
fundamental combustion experiments.\footnote{ Interestingly, this contrasts the
situation for chemical kinetic models, where the CHEMKIN~\autocite{Kee:1996ck}
format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or
FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code} lag behind considerably,
although use of the former continually grows due to Cantera's open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?
}

Instead, such data is typically found in under-documented comma-separated value
files and Excel spreadsheets, or contained in PDF tables rather than as
supplementary material associated with a paper. In the worst case, data is only
available in figures and must be digitized by the use of software such as
WebPlotDigitizer~\autocite{WebPlotDigitizer}. These practices limit the wide use
of the valuable, extensive experimental data available in the literature.
Frenklach~\autocite{Frenklach:2007bm} further explained the benefits of
standardized, widely available data for combustion. In brief, much of
fundamental combustion\slash chemical kinetics research drives towards the
ultimate goal of developing predictive kinetic models---and ultimately this
depends on a \emph{community} infrastructure for data and methods.

The combustion community has not yet widely adopted a standard format for
experimental data, but some researchers and groups have proposed solutions to
this problem. Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess
\textbf{I}nformatics \textbf{M}od\textbf{e}l) data
format~\autocite{Frenklach:2007bm,You:2011hy} and an associated online database,
the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated
quantities, using the Extensible Markup Language (XML) standard.

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make
it a strong standardization format, but it suffers from several flaws that
prevent wide adoption. First, the PrIMe standard does not require or support all
information needed to simulate an experiment, including a machine-readable
definition of ignition or a standard way to express detailed facility specific
effects important to properly simulating certain experiments. Second, the XML
format is intended to be a markup language rather than a data format, and
presents a barrier to creating and working with database files. Furthermore, the
focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for
scholarly products and InCHI or SMILES for species), limits usability outside
the PrIMe ecosystem. Finally, the closed and opaque nature of the PrIMe standard
and associated Data Warehouse limit contributions and community development of
the PrIMe standard and compatible data files and tools.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the
ReSpecTh standard, which builds on the PrIMe format. ReSpecTh adds important
features that make files better standalone representations of experimental
data---i.e., more informative by themselves rather than in concert with a larger
system. For example, where PrIMe uses internal bibliographic references,
ReSpecTh adds a field for typical bibliographic data. The ReSpecTh standard also
provides machine-readable formats for specifying ignition experiments, including
a field for the definition of autoignition and the ability to specify
facility-specific effects. However, ReSpecTh experimental files are another
XML-based format, and as such, suffer from the same usability issues as PrIMe.
Moreover, the standard remains closed, requiring potential users to register
with the standard's authors to access the specification.

Nonetheless, the usefulness of databases of fundamental combustion experiments
has been shown to be quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen\slash
syngas~\autocite{Varga:2016gj} and ethanol~\autocite{Olm:2016et}. In all four
cases, they converted numerous experimental datasets from the literature into
the ReSpecTh standard.

Despite the development of these standards, limited examples of open data
\emph{sharing} utilizing the formats is available in the literature. To the best
of our knowledge, the largest such such publicly available database is hosted by
the Clean Combustion Research Center at King Abdullah University of Science and
Technology (KAUST) and is known as CloudFlame
(\url{https://cloudflame.kaust.edu.sa}). Started in
2013~\autocite{Goteng:2013cf},\todo{a few more refs needed}{} CloudFlame serves
as an openly accessible database for experimental data, available in a standard
CSV format, and also provides a cloud infrastructure for running simulations
based on stored models and data. While an admirable and useful effort,
CloudFlame, like ReSpecTh and PrIMe, is a closed system controlled by a single
institution rather than by the community at large.

Therefore, we believe there is a need for an open, community focused and
developed, combustion\slash chemical kinetics data format. In this work, we
present a new, open-source, human- and machine-readable data standard for
fundamental combustion experiments, \ck{}, and offer tools for easily working
with data encoded in this format. Niemeyer recently introduced an initial
version of the \ck{} format~\autocite{Niemeyer:2016wf}, and in this work we
further formalize and develop the standard. We also discuss a related Python
software tool for validating and working with \ck{} files: \pk{}. Our
motivations are similar those of the PrIMe and ReSpecTh teams: the \ck{} project
will enable easy sharing and use of fundamental combustion data, for the
(primary) purposes of developing and validating predictive chemical kinetic
models. However, we make usability a major design focus, and plan to share all
data and software openly to cultivate a community of user-contributors.

@@

\ck{} and \pk{} are currently under active development, and contributions from
the community are welcomed. All development occurs under the BSD
3-clause license and the code is housed on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}. Future directions for development
are outlined in the public
\href{Roadmap}{https://github.com/pr-omethe-us/PyKED/wiki/Roadmap}. The highest
priority issues are currently adding support for other types of fundamental
experiments, including flame speed, speciation, and flame extinction
measurements.
